### 2021-07-14 response.tex

\rev{Reviewer \#1}

\revtext{This is a review of ``Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants'' by Park and others for Royal Society Interface.  This is the latest in an ongoing series of papers (Champredon 2015, Champredon 2018, Park 2019, Dushoff 2021, ...) in which the last author has considered aspects of incubation, growth, and reproduction in SIR epidemic theory.  In this manuscript, the authors use theory to anticipate the effects of COVID-19 variants where the distribution of infectivity evolves.}

\revtext{The manuscript is excellently written. The figures are clear.  All the mathematics seems satisfactory.  Unfortunately, the manuscript does not make its case for novel results.}

\jd{If we said Thank you a little less, it might mean a little more.}

Thank you for the review.
We have tried to frame our results better, and to clarify both the novelty and importance of this study.
We agree with Reviewer 3 that, while many of these concepts have been raised in the context of the recent emergence of new SARS-CoV-2 variants, careful treatment has been lacking. 
We believe that the addition of a new simulations (Figure 5) and the revised text better highlight the novelty of our work.

@@

\revtext{The title is overly-broad as it suggests consideration of general forms of generation interval distributions (along the lines of Miller 2010 10.1016/j.jtbi.2009.08.007), when infect the authors study variation around the time of peak infectivity using a (very standard) $\gamma$-chain trick.  The study of generation time and $\gamma$ chains has a long history from the cited-but-unreviewed Svensson 2007, Wallinga 2007, to uncited Kenah 2008, A.  Lloyd 2001, D. Anderson 1980, Bailey 1964 (see Park 2019 and Champredon 2015).  In light of these and other works, the title and abstract fail to identify novel contributions to our understanding of epidemic theory.}

\swp{JD, we need a new title.} \jd{I think it helps a lot to just not say ``distributions'' since we're mostly focused on changes in the mean.}

\jd{Have we looked at the Miller paper?}

Thank you for the suggestions. We have changed the title and the abstract to highlight the novelty of our work. See also below.

@@

``Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).''

Assuming Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We [now?] explain this later in Secion 2:

@@

\revtext{Other comments:}

\revtext{Topic on which there is a considerable confusion in terminology in the literature.  In addition to "generation interval", terms "generation time", "transmission interval", "serial interval", and others have been used in the literature, all with related by different definitions.  This manuscripts definition "time between infection and transmission" implies a single transmission event, but Kenah pointed out that time-contraction appears when allowing multiple transmission events (see Champredon 2015).  A clearer definition up front will help readers.  Fortunately, the results don't depend on the specific definition.}

Thank you for the suggestions. We have tried to clarify what we mean by the generation interval. In particular, often-menetioned time contractions of generation intervals apply to the forward realized generation interval, which measure time between actual infection events, rather than the intrinsic generation interval, which measure the intrinsic infectiousness of infected individuals. Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts. These distinctions are explained in Champredon 2015, and we have tried to clarify these ideas again here in Section 2:

@@

\revtext{The 'speed'-'strength' distinction from the previous paper is seductive, but perhaps leaning into general quantitative illiteracy.  When one pays attention to units -- "speed" L/T,  "Growth rate" 1/T while "strength" has dimensions of force, typically, but reproduction number is a dimensionless ratio of cases per case.  Beyond time-independent final-size relations (which are themselves not robust to effects like behavior change), several researchers have argued that growth rates generally more useful than reproduction numbers (giving equivalent threshold conditions, for example) and not as prone to mis-interpretation.  Other authors in evolutionary biology and economics have argued that the reproduction number's scientific value is best salvaged using the "discounted reproduction success" (McNamara, Houston, and Collins 2001), rather than fighting of an R-r straw-man.}

We are certainly not trying to lean in to innumeracy. We agree that ``strength'' and ``speed'' are not used in the classical physics sense -- instead, we try to be clear how we define them (and have tried to clarify further).

We have argued (https://doi.org/10.1098/rspb.2020.1556) that large segments of the epidemiology and modeling community have an over-emphasis on R, and have tried to use the terms speed and strength to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on control).
As we argue here and elsewhere, we believe that sometimes one perspective ro another gives a clearer picture.
For example, in the beginning of an epidemic, it is more appropriate to fix epidemic speed $r$ when evaluating uncertainties in the generation-interval distribution.
On the other hand, it is more often appropriate to fix relative strength $\rho$, rather than relative speed $\delta$, when conditions are changing due to interventions.
We hope that our revisions, including the addition of a new figure (see response to reviewer 3), have made these points clearer.
We thus decided to keep the terms ``speed'' and ``strength'' in this manuscript.

@@

\revtext{Equation 1 As Kermack and McKendrick introduced it, the compartmental model is a convenient *specialization* of the integral models-- see the nice re-analysis by Breda et al., 2012.}

Thank you for pointing this out. We now cite Breda et al., and acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''

@@

``When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.''

\revtext{p 4 l55.  Shape of the distribution matters -- heavy tails and failed central concentration.  This detail will escape the naive reader unless we poke them a little more with it.}

Thank you for the suggestion. We have reworded this section:

\jd{We should deal separately with $\kappa$ and with the gamma-generation assumption.
``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.''

@@

\rev{Reviewer \#2}

\revtext{This is a well-written paper about the role of the generation interval distribution in the estimation of the relative strength (reproduction number) and speed
(epidemic growth rate) of a new variant of a disease. Using a renewal equation
framework, they consider inferring relative strength from relative speed and vice
versa as well as inferring relative strength from incidence data. While all of the
analysis in the paper is sound, I am concerned that the strong assumptions made
(mass-action epidemic, constant generation interval distribution, etc.) will limit
the reliability of the results in a practical application. However, they clearly
establish that assumptions about the generation interval distribution have important implications for the analysis of novel variants. While I am generally
skeptical of population-level analyses of infectious disease transmission based
on generation intervals, these are done frequently so this is a useful insight.
More detailed comments are below.}

Thank you for the review. We have tried to make the limitation and assumptions clearer throughout the text. \jd{Anything about the generality? I don't feel we rely much on mass-action or constant distributions.}

@@

\revtext{1. (page 3, first paragraph) It is important to acknowledge that the renewal equation approach to the relationship between the generation interval
distribution, R, and the growth rate r assumes a mass-action epidemic.
When we assume a fixed generation interval distribution, we must add the
assumption that the rate of depletion of susceptibles over time is negligible and the prevalence of infection is low. One of the most important
findings of Ref 10 is that the mean generation interval contracts when the
rate of depletion of susceptibles increases. While the renewal equation
does allow for the generation interval to change over time, the analyses
in Sections 3–5 assume fixed mean generation intervals. Because variants
are likely to spread when susceptibles are being depleted by both variants
(due to cross-protection) and when the prevalence of infection is not low, these implicit assumptions place significant limitations on the practical
implications of the analysis.}

Thank you for pointing this out.
We acknowledge that we are assuming mass action.
However, a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
\jd{Can we say something about the rR link at any particular time? Doesn't that also work?}
This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
We have added a paragraph in Section 2 to clarify this distinction (please see response to reviewer 1 for text).

@@

\revtext{(page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
speed-like interventions work by identifying potential infections quickly
(effectively ending the infectious period early while not necessarily altering
the hazard of infectious contact) and strength-like interventions work by
reducing the hazard of infectious contact (while not necessarily altering
the duration of infectiousness). A good explanation of these ideas is on
lines 228–231 on page 10.}

Thank you for the suggestion. We now define them more clearly throughout the text. \jd{Presumably there's a particular place where we lay out the definitions.}

@@

\revtext{Abstract: 37: On first reading, the distinction between 'speed-like' and 'strength-like' interventions wasn't clear to me, nor why contact tracing should target r but not R, and why social distancing should target R but not r. Could you use more recognizable terms here to highlight how these interventions differ from one another? (is the key difference that contact tracing reduces the number of infectious individuals and thus the probability that a susceptible person comes into contact with an infectious person, while social distancing reduces the per-contact chance of transmission? Or something else?) This comes up again in line 57 of the Introduction.}

Thank you for pointing this out. We have tried to make these distinctions clearer throughout text.

\revtext{166: I think it would be worth stating the range of Rwt and/or Rvar that yields this range for delta (I think it's something like 0.49-1.27 for the wild type, 0.78-2.04 for the variant, based on Figure 2 - but would be helpful if the reader didn't have to derive). Is the varying R what you're referring to when you speak of "underlying epidemiological conditions" in 166-167?}

Thank you for this suggestion. We now clarify this when it is mentioned for the first time:

\jd{This is the sort of text I was hinting at above about mass action.}
``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.''

### 2021-07-14 newvariant.tex
Once through the response, and some related changes in the MS.

\begin{flushleft}{
	\Large
	\textbf\newline{
	  Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Benjamin M.\ Bolker\textsuperscript{2,3,4}
Sebastian Funk\textsuperscript{5,6}
C.\ Jessica E.\ Metcalf\textsuperscript{1,7}
Joshua S.\ Weitz\textsuperscript{8,9}
Bryan T.\ Grenfell\textsuperscript{1,7}
Jonathan Dushoff\textsuperscript{2,3,4}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} Department for Infectious Disease Epidemiology, London School of Hygiene and Tropical Medicine, London, UK
\\
\textbf{6} Centre for Mathematical Modelling of Infectious Diseases, London School of Hygiene and Tropical Medicine, London, UK
\\
\textbf{7} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\textbf{8} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\bigskip

@@

Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
We further argue that holding $\rho$ constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how $\rho$ and $\delta$ are affected by two idealized interventions.
A constant-strength intervention, which reduces transmission rates by a constant amount, affects $\delta$ but not $\rho$---this is analogous to vaccination and social distancing.
A constant-speed intervention, which isolates infected individuals at a constant hazard, affects $\rho$ but not $\delta$---this is analogous to contact tracing.
Uncertainties in the generation-interval distribution determine the effectiveness of these interventions.

@@

Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection. 
Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
Epidemic ``speed'' is characterized by the growth rate $r$, which has a unit of $1/\mathrm{time}$ and describes how fast a disease spreads at the population level.
Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.

Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
Instead, epidemic strength can be inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).

@@

Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.

@@

Neglecting the (relatively slow) rate of new mutations and assuming a mass-action epidemic, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.

@@

Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
% The 
Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
	  The importance of generation intervals in investigating dynamics and control of new SARS-CoV-2 variants }
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Benjamin M.\ Bolker\textsuperscript{2,3,4}
Sebastian Funk\textsuperscript{5,6}
C.\ Jessica E.\ Metcalf\textsuperscript{1,7}
Joshua S.\ Weitz\textsuperscript{8,9}
Bryan T.\ Grenfell\textsuperscript{1,7}
Jonathan Dushoff\textsuperscript{2,3,4}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} Department for Infectious Disease Epidemiology, London School of Hygiene and Tropical Medicine, London, UK
\\
\textbf{6} Centre for Mathematical Modelling of Infectious Diseases, London School of Hygiene and Tropical Medicine, London, UK
\\
\textbf{7} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\textbf{8} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\bigskip

@@

Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the alpha \jd{suggesting that we use spelled-out greek names for variants, to reduce confusion with rendered greek letters for quantities} variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
We further argue that holding $\rho$ constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how $\rho$ and $\delta$ are affected by two idealized interventions.
A constant-strength intervention, which reduces transmission rates by a constant amount, affects $\delta$ but not $\rho$---this is analogous to vaccination and social distancing.
A constant-speed intervention, which isolates infected individuals at a constant hazard, affects $\rho$ but not $\delta$---this is analogous to contact tracing.
Uncertainties in the generation-interval distribution determine the effectiveness of these interventions.

@@

Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection. 
Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of disease spreads at the population level.
Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.

Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
\jd{``instead'' seems wrong here, since we cite davies2021estimated in both lists. What's going on?}
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.

For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
In this same case, however, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.

Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with strength-like interventions (but not with speed-like interventions, whose effectiveness is fixed if speed is fixed).

@@

Here, we explore how different assumptions affect estimates of the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$) for a new variant, such as B.1.1.7 \jd{search and replace variant names if you agree}.
% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.

@@

Neglecting the (relatively slow) rate of new mutations and assuming a mass-action epidemic, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.

@@

\jd{Where did -based come from? Do we like it?}
Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}--- even though the instantaneous generation-interval distribution remains unchanged in this scenario.
While the instantaneous generation-interval distribution can change under speed-based changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
% The 
Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).
